\documentclass[sigplan,screen]{acmart}

\PassOptionsToPackage{activate=false,babel=false}{microtype}
\usepackage{mathpartir}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{xspace}
\usepackage{tikz}
\usetikzlibrary{decorations.pathreplacing,arrows.meta,positioning}

%% ---- Custom commands ----
%% Reference to a Lean 4 artifact
\newcommand{\leanref}[1]{\texttt{#1}}

%% Multi-letter math identifiers
\newcommand{\YjsLt}{\mathrel{<_Y}}
\newcommand{\YjsLeq}{\mathrel{\leq_Y}}
\newcommand{\hb}{\mathrel{\prec}}
\newcommand{\hble}{\mathrel{\preceq}}
\newcommand{\CC}{\mathit{CC}}

%% Semantic domains
\newcommand{\First}{\mathsf{first}}
\newcommand{\Last}{\mathsf{last}}
\newcommand{\OriginLt}{\mathrel{<_{\mathit{orig}}}}

%% BNF helpers
\newcommand{\bnfdef}{\;::=\;}
\newcommand{\bnfmid}{\;\mid\;}

%% Named sets / functions
\newcommand{\interpOps}{\mathit{interpOps}}
\newcommand{\effect}{\mathit{effect}}
\newcommand{\init}{\mathit{init}}
\newcommand{\histories}{\mathcal{H}}
\newcommand{\Deliver}{\mathsf{Deliver}}
\newcommand{\Broadcast}{\mathsf{Broadcast}}
\newcommand{\Delivered}[1]{\mathit{delivered}_{#1}}

\copyrightyear{2025}
\acmYear{2025}

\author{Izumi Asakura}
\email{i.asakura58@gmail.com}

\begin{document}

\title{Conditional Commutativity: A Generic Convergence Proof for
  CRDTs and Its Application to Yjs}

\begin{abstract}
A standard technique for proving convergence of operation-based CRDTs is to
show that concurrent operations commute on \emph{all} states.  We demonstrate
that this unconditional commutativity fails for Yjs, one of the most widely
deployed collaborative editing frameworks, and exhibit a mechanized
counterexample in Lean~4.  To close this gap, we introduce a
\emph{conditional commutativity} framework that requires commutativity only
on states satisfying an invariant maintained by causal delivery.  We prove a
generic convergence theorem within this framework and instantiate it for Yjs,
obtaining the first complete machine-checked proof that the Yjs core CRDT
algorithm achieves strong eventual consistency.  The instantiation establishes
that Yjs insert and delete operations preserve a strict total order on items
and rests on a transport lemma carrying broadcast-time item validity to any
causally compatible state.  The entire development is formalized in Lean~4
with no axioms beyond those of its type theory.
\end{abstract}

\maketitle

%%
%% -------------------------------------------------------------------
\section{Introduction}
\label{sec:intro}
%%

%% P1: Motivation
Yjs~\cite{Yjs2023} is one of the most widely deployed conflict-free replicated
data type (CRDT) implementations, powering collaborative editors such as
TipTap.  Despite its practical importance, Yjs lacks a complete correctness
proof.  The YATA paper~\cite{Nicolaescu2016}, which describes the algorithm
underlying Yjs, provides only partial proofs for selected lemmas, and we
later discovered a bug in the original algorithm that required a fix in the
Yjs implementation~\cite{YjsBugDiscussion2024}.

%% P3: Standard proof pattern — unconditional commutativity
A standard proof technique for operation-based CRDTs~\cite{Shapiro2011} is
\emph{unconditional commutativity}: for all operations $x$, $y$ and
every state~$s$,
$\effect(x, \effect(y, s)) = \effect(y, \effect(x, s))$.
Gomes et al.~\cite{Gomes2017} formalized this approach in Isabelle/HOL
and showed that unconditional commutativity, together with causal
delivery, implies convergence.

%% P4: Why it fails for Yjs
For Yjs, unconditional commutativity does not hold.  We exhibit a
mechanized counterexample (\S\ref{sec:overview}): two concurrent inserts
produce different results depending on delivery order.  The root cause is
that in Yjs, the \emph{validity} of an insert operation depends on the
state to which it is applied---an item's origin and right-origin must
satisfy a structural invariant relative to the existing items.  When the
effect function is applied to a state where the operation is invalid, the
commutativity equation can fail.  Since unconditional commutativity
quantifies over \emph{all} states, such invalid combinations are enough
to break the property entirely.

%% P5: Our solution — conditional commutativity
We address this with \emph{conditional commutativity}: for all operations
$x$, $y$ and every state~$s$,
if $\mathit{Valid}(x, s)$ and $\mathit{Valid}(y, s)$, then
$\effect(x, \effect(y, s)) = \effect(y, \effect(x, s))$.  This alone is
not enough---the convergence proof reorders operations, so an operation
may be applied against a different state than the one it was broadcast
against.  We therefore require a second condition, \emph{causal validity}:
each broadcast operation remains valid at every state produced by an
$\hb$-consistent execution containing its causal predecessors.  Together,
conditional commutativity and causal validity yield a generic convergence
theorem (Theorem~\ref{thm:generic-conv}) that is independent of any
particular CRDT.  We instantiate it for Yjs, obtaining the first complete
machine-checked proof of strong eventual consistency for the Yjs core
CRDT algorithm (Theorem~\ref{thm:yjs-convergence}).

%% P6: Contributions
This work makes two contributions.
\begin{enumerate}
\item \textbf{A generic conditional commutativity convergence framework.}
  We identify that unconditional concurrent commutativity fails for Yjs
  (exhibited as a mechanized counterexample) and, to overcome this, we
  generalize the standard commutativity-based convergence argument to require
  commutativity only on invariant-satisfying states, formalized as a reusable
  theorem in Lean~4.  The proof establishes convergence by reordering the
  delivery list through repeated swaps of adjacent concurrent operations.
\item \textbf{First formal verification of the Yjs core CRDT algorithm.}
  We instantiate the generic framework for Yjs, providing the first
  machine-checked proof that the core algorithm achieves strong eventual
  consistency.  Our main lemma is the preservation of the total order on items
  by the Yjs insert and delete operations.  In particular, the insert
  operation is defined by a non-trivial procedural loop, and establishing its
  order-preserving behavior requires discovering a four-part invariant
  maintained by the scan.  We also mechanize a strict total order on Yjs
  items---with totality, transitivity, and asymmetry---together with a
  no-cross-origin lemma that drives the preservation proof; the inference
  rules of the order are carefully balanced so that transitivity and asymmetry
  can be recovered via size-based induction.
\end{enumerate}

%% P7: Paper organization
Section~\ref{sec:definitions} defines the network model and Yjs data
structures.
Section~\ref{sec:overview} presents the counterexample motivating
conditional commutativity and gives an overview of the proof architecture.
Sections~\ref{sec:order}--\ref{sec:convergence} develop the proof.
Section~\ref{sec:related} discusses related work, and
Section~\ref{sec:conclusion} concludes.
Detailed proofs for the order-theoretic properties appear in
Appendix~\ref{app:order-proofs}.

%%
%% -------------------------------------------------------------------
\section{Network and Execution Model}
\label{sec:definitions}
%%

This section formalizes the setting in which Yjs operates.  We first fix a
generic operation network model, following Gomes et al.~\cite{Gomes2017}, that
captures operation-based CRDT executions under causal delivery.  We then
specialize it with Yjs-specific data structures and state the main convergence
theorem.

%%
\subsection{Network Model}
\label{subsec:network}
%%

Our goal in this subsection is to formalize what it means for a collection of
replicas to execute an operation-based CRDT.  A network consists of several
replicas indexed by a set $N$; each replica maintains a local sequence of
events that records, in order, the \emph{broadcast} of a locally issued
operation and the \emph{deliver} events at which operations (either local or
remote) are applied to the local state.  A broadcast event disseminates the
operation to other replicas, whereas a deliver event is the single site at
which operations actually modify the state.  On top of this bare structure, we
add a causality condition (so the dependency order of broadcasts is respected
on delivery) and then equip operations with a transition semantics.  The
resulting model is a mild reformulation of the OOPSLA'17 framework of
Gomes~et~al.~\cite{Gomes2017}.

%% P1: Messages, Events, Node Histories, Local Ordering
The model is parameterized by a message type~$A$.
Each message $m \in A$ carries an identifier $\mathit{msgId}(m) \in M$
that uniquely determines it.

\begin{definition}[Event (\leanref{Event})]\label{def:event}
An \emph{event} is either the broadcast or delivery of a message:
\[
  e \bnfdef \Broadcast(m) \bnfmid \Deliver(m)
\]
\end{definition}

\begin{definition}[Node history (\leanref{NodeHistories})]\label{def:node-history}
Fix a set $N$ of node identifiers.  A \emph{node history} $H_i$ for node
$i \in N$ is a duplicate-free list of events.  For events $e_1, e_2 \in H_i$
we write $e_1 <_i e_2$ (\emph{locally ordered}) when $e_1$ appears before
$e_2$ in $H_i$.
\end{definition}

%% P2: Network, Happens-Before, Causal Network
\begin{definition}[Network (\leanref{NetworkBase})]\label{def:network-base}
A \emph{network} is a collection of node histories
$\{H_i \mid i \in N\}$ satisfying the following properties:
\begin{itemize}
\item[(N1)] Every delivered message has a broadcasting node:
  $\Deliver(m) \in H_i \Rightarrow \exists j.\; \Broadcast(m) \in H_j$.
\item[(N2)] Broadcast precedes delivery locally:
  $\Deliver(m) \in H_i \Rightarrow \Broadcast(m) <_i \Deliver(m)$.
\item[(N3)] Message IDs are globally unique:
  $\mathit{msgId}(m_1) = \mathit{msgId}(m_2)$ and both broadcast
  $\Rightarrow m_1 = m_2$.
\end{itemize}
\end{definition}

Operation-based CRDTs rely on a \emph{causal network}: the network is required
to deliver messages in an order that respects the happens-before relation of
Lamport~\cite{Lamport1978}, so that every message encounters its causal
predecessors at replay time.  We make this precise next.

\begin{definition}[Happens-before (\leanref{HappensBefore})]\label{def:happens-before}
The \emph{happens-before} relation $\hb$ is the least relation satisfying:
\begin{mathpar}
\inferrule*[right=HB-Local-Bc]
  {\Broadcast(m_1) <_i \Broadcast(m_2)}
  {m_1 \hb m_2}
\and
\inferrule*[right=HB-Deliver]
  {\Deliver(m_1) <_i \Broadcast(m_2)}
  {m_1 \hb m_2}
\and
\inferrule*[right=HB-Trans]
  {m_1 \hb m_2 \\ m_2 \hb m_3}
  {m_1 \hb m_3}
\end{mathpar}
\end{definition}

\begin{definition}[Causal network (\leanref{CausalNetwork})]\label{def:causal-network}
A \emph{causal network} extends a network with the property that if
$m_2$ is delivered at node~$i$ and $m_1 \hb m_2$, then $\Deliver(m_1) <_i
\Deliver(m_2)$.
\end{definition}

%% P3: Operation, Operation Network, Delivered Messages, Interpretation
So far, the message type~$A$ only carries an identifier.
An \emph{operation network} equips messages with additional structure---a
state-transition semantics and a validity predicate---so that messages can be
interpreted as state-transforming operations.  We first define the
state-transition structure, then the delivered-message projection that turns a
history into the sequence of operations that actually affect state, and
finally the validity predicate.

\begin{definition}[Operation (\leanref{Operation})]\label{def:operation}
An \emph{operation structure} on a message type~$A$ is a tuple
$(\mathit{St},\; \init,\; \effect)$ where
$\mathit{St}$ is the set of states,
$\init \in \mathit{St}$ is the initial state, and
$\effect : A \to \mathit{St} \to \mathit{St}_\bot$ is a partial effect
function; here $\mathit{St}_\bot = \mathit{St} + \{\bot\}$ denotes the state
type extended with a single failure value, written $\bot$.\footnote{In
the Lean~4 artifact the failure value carries a string error message
(\leanref{Err}); we omit this detail here because the error payload plays no
role in the proofs.}
\end{definition}
\noindent
We write $\interpOps(ms, s)$ for the result of folding $\effect$ over a message
list~$ms$ starting from state~$s$.

\begin{definition}[Delivered messages]\label{def:delivered}
The \emph{delivered messages} at node~$i$, written
$\Delivered{i}(H_i)$, is the sublist of $H_i$ consisting of all
$\Deliver$ events projected to their message payloads.
\end{definition}
\noindent
Because broadcasts have no effect, the local state at node~$i$ after running
the history $H_i$ is precisely
$\interpOps(\Delivered{i}(H_i),\, \init)$.

\begin{definition}[Operation network (\leanref{OperationNetwork})]\label{def:operation-network}
An \emph{operation network} extends a causal network whose message type~$A$
is equipped with both an operation structure
(Definition~\ref{def:operation}) and a validity predicate
$\mathit{isValid} : \mathit{St} \to A \to \mathit{Prop}$, subject to the
following condition.
\begin{itemize}
\item[(V)] \emph{Broadcast validity.} For every node history $H_i$ with
$H_i = \mathit{pre} \mathbin{+\!\!+}
[\Broadcast(m)] \mathbin{+\!\!+} \mathit{post}$, there exists a state $s$
such that $\interpOps(\Delivered{i}(\mathit{pre}), \init) = s$ and
$\mathit{isValid}(s, m)$ holds.
\end{itemize}
\end{definition}
\noindent
Condition~(V) is the OOPSLA'17 valid-message property~\cite{Gomes2017}: every
locally broadcast message must be valid at the local state reached just before
the broadcast.

%%
\subsection{Yjs Data Model}
\label{subsec:yjs-data}
%%

We now define the data structures that represent a Yjs document.
A document is a sequence of \emph{items}, each recording the neighbors it was
inserted between (its \emph{origin} and \emph{right-origin}).  This subsection
fixes those data structures; we make no attempt to model garbage collection,
undo/redo, subdocuments, or the awareness protocol of the Yjs
manual~\cite{Yjs2023}.

%% P4: YjsId, YjsPtr, YjsItem
\begin{definition}[Yjs identifier (\leanref{YjsId})]\label{def:yjs-id}
A \emph{Yjs identifier} pairs a client identifier with a logical clock:
$\mathit{YjsId} = \mathit{ClientId} \times \mathbb{N}$.
Identifiers are compared lexicographically.
\end{definition}

Each item records who its immediate left and right neighbors were at the
moment of the originating broadcast.  Because the endpoints of a document
have no such neighbors, we introduce two sentinel pointers, $\First$ and
$\Last$, so that insertions at either end still have well-defined origins.
A pointer is therefore either one of these sentinels or a reference to
another item; in particular, pointers and items are mutually recursive.

\begin{definition}[Yjs pointer and item (\leanref{YjsPtr}, \leanref{YjsItem})]
\label{def:yjs-item}
\emph{Yjs pointers}~$p$ and \emph{Yjs items}~$x$ are defined by mutual
recursion:
\begin{align*}
  p &\bnfdef \First \bnfmid \Last \bnfmid \langle x \rangle \\
  x &\bnfdef \{
    \mathit{origin} : p,\;
    \mathit{rightOrigin} : p,\;
    \mathit{id} : \mathit{YjsId},\;
    \mathit{content} : A \}
\end{align*}
The $\mathit{origin}$ and $\mathit{rightOrigin}$ fields store the left and
right neighbors the item was inserted between at broadcast time, using
$\First$ or $\Last$ as sentinels when the item is inserted at a document
boundary.\footnote{The Lean~4 formalization embeds items directly inside
pointers, making origins concrete sub-terms; the real Yjs implementation
instead references origins by identifier.  The artifact contains an
alternative \emph{indirect} variant (\texttt{LeanYjs/Indirect}) that follows
the implementation faithfully, though its verification is not complete.}
\end{definition}

%% P5: YjsState, deletion
\begin{definition}[Yjs state (\leanref{YjsState})]\label{def:yjs-state}
A \emph{Yjs state} consists of an array of all items inserted so far
(including those that have since been deleted) together with a set of
deleted identifiers:
\[
  \mathit{YjsState}(A) = \{
    \mathit{items} : \mathit{Array}(\mathit{YjsItem}(A)),\;
    \mathit{deletedIds} : \mathit{Set}(\mathit{YjsId}) \}
\]
Deletion is implemented logically via tombstones: the deleted identifier is
added to $\mathit{deletedIds}$ while the item remains in the array, preserving
the anchor points needed by subsequent insertions.
\end{definition}

%%
\subsection{Yjs as an Operation Network Instance}
\label{subsec:yjs-network}
%%

To specialize the generic operation network model of
\S\ref{subsec:network} to Yjs, we give the message type, its operation
structure, and the validity predicate; we additionally impose two
Yjs-specific well-formedness properties on how clients issue broadcasts.

%% P6: YjsOperation, Effect, Insert Algorithm, YjsOperationNetwork
\begin{definition}[Yjs operation (\leanref{YjsOperation})]\label{def:yjs-operation}
A \emph{Yjs operation} either inserts a new item or marks an existing one as
deleted:
\begin{align*}
  \mathit{op} &\bnfdef \mathsf{Insert}(\mathit{inp})
                \bnfmid \mathsf{Delete}(\mathit{id},\, \mathit{deletedId}) \\
  \mathit{inp} &\bnfdef \{
    \mathit{originId} : \mathit{YjsId},\;
    \mathit{rightOriginId} : \mathit{YjsId},\; \\
  &\qquad\; \mathit{content} : A,\;
    \mathit{id} : \mathit{YjsId} \}
\end{align*}
where $\mathit{inp}$ is an \emph{insert input} recording the identifiers of
the origin and right-origin neighbors, the content, and the new item's
identifier.  This instantiates Definition~\ref{def:operation} with
$\mathit{St} = \mathit{YjsState}$,
$\mathit{init} = ([\,],\; \emptyset)$, and
\[
  \effect(\mathit{op},\, \sigma) \;=\;
  \begin{cases}
    \mathit{insert}(\sigma,\, \mathit{inp})
      & \text{if } \mathit{op} = \mathsf{Insert}(\mathit{inp}) \\
    \mathit{deleteById}(\sigma,\, \mathit{did})
      & \text{if } \mathit{op} = \mathsf{Delete}(\mathit{id},\, \mathit{did})
  \end{cases}
\]
Deletion simply adds $\mathit{did}$ to the tombstone set, so it always
succeeds.  Insertion is defined by Algorithm~\ref{alg:integrate}: starting
from the origin, the scan walks rightward through the existing items,
comparing each encountered item against the new one using the order
developed in \S\ref{sec:order}, and inserts the new item at the first
position where no conflicting item precedes it.  The detailed invariant
maintained by the loop is analyzed in \S\ref{sec:algorithm}.
\end{definition}

\begin{algorithm}[t]
\caption{Yjs insert algorithm}\label{alg:integrate}
\begin{algorithmic}[1]
\Require{insert input $\mathit{inp}$, items array $\mathit{arr}$}
\State $\mathit{leftIdx} \gets \mathit{findIndex}(\mathit{inp.originId},\, \mathit{arr})$
\Comment{$-1$ if origin is $\First$}
\State $\mathit{rightIdx} \gets \mathit{findIndex}(\mathit{inp.rightOriginId},\, \mathit{arr})$
\Comment{$|\mathit{arr}|$ if right-origin is $\Last$}
\State $\mathit{dest} \gets \mathit{leftIdx} + 1$
\State $\mathit{scanning} \gets \mathbf{false}$
\For{$i \gets \mathit{leftIdx}+1$ \textbf{to} $\mathit{rightIdx}-1$}
  \State $\mathit{other} \gets \mathit{arr}[i]$
  \State $\mathit{oLeft} \gets \mathit{findIndex}(\mathit{other.origin},\, \mathit{arr})$
  \State $\mathit{oRight} \gets \mathit{findIndex}(\mathit{other.rightOrigin},\, \mathit{arr})$
  \If{$\mathit{oLeft} < \mathit{leftIdx}$}
    \State \textbf{break}
  \ElsIf{$\mathit{oLeft} = \mathit{leftIdx}$}
    \If{$\mathit{inp.clientId} > \mathit{other.clientId}$}
      \State $\mathit{scanning} \gets \mathbf{false}$
    \ElsIf{$\mathit{oRight} = \mathit{rightIdx}$}
      \State \textbf{break}
    \Else
      \State $\mathit{scanning} \gets \mathbf{true}$
    \EndIf
  \EndIf
  \If{$\neg\, \mathit{scanning}$}
    \State $\mathit{dest} \gets i + 1$
  \EndIf
\EndFor
\State \Return $\mathit{arr}.\mathit{insert}(\mathit{dest},\, \mathit{mkItem}(\mathit{inp}, \mathit{arr}))$
\end{algorithmic}
\end{algorithm}

\begin{definition}[Yjs operation network (\leanref{YjsOperationNetwork})]\label{def:yjs-op-network}
A \emph{Yjs operation network} is an operation network on the Yjs operation
type satisfying the following Yjs-specific properties on broadcasts:
\begin{itemize}
\item[(Y1)] \textbf{Client ID consistency.} Every operation broadcast by node
  $i$ carries client ID $i$.
\item[(Y2)] \textbf{Clock monotonicity.} The clock component of the
  identifiers broadcast by a single client is strictly increasing along the
  local history.
\end{itemize}
Both properties hold in any faithful Yjs implementation; they ensure, in
particular, that item identifiers are globally unique.
\end{definition}

%%
\subsection{Main Theorem}
\label{subsec:main-theorem}
%%

%% P7
We can now state the main convergence result.

\begin{theorem}[Yjs Convergence, \leanref{YjsOperationNetwork\_converge'}]
\label{thm:convergence}
For any Yjs operation network and any two nodes $i$, $j$, if
$\Delivered{i}(H_i) \approx \Delivered{j}(H_j)$ (equality as sets), then
$\interpOps(\Delivered{i}(H_i), \init) =
 \interpOps(\Delivered{j}(H_j), \init)$.
\end{theorem}

\noindent
Here $\approx$ denotes that the two lists contain the same set of messages;
the ordering may differ.  That is, any two nodes that have delivered the same
set of operations converge to equal states.

%%
%% -------------------------------------------------------------------
\section{Overview}
\label{sec:overview}
%%

This section motivates the shape of our proof.  We first demonstrate that
the standard, unconditional commutativity strategy cannot be used for Yjs
by exhibiting a concrete counterexample; we then sketch how a conditional
commutativity argument recovers convergence and indicate how the remaining
sections fit together.

\subsection{Counterexample: Unconditional Commutativity Fails}

We exhibit a mechanized counterexample
(\leanref{inconsistent1\_inconsistent2}) showing that two concurrent Yjs
insert operations can produce different results depending on delivery order.

\smallskip\noindent\textbf{Initial state.}
Consider a document \texttt{"ab"} consisting of two items produced by two
earlier (concurrent) insert operations:
$a$ (content \texttt{"a"}, id $(0,0)$, origin $\First$, right-origin $\Last$)
and $b$ (content \texttt{"b"}, id $(1,0)$, origin $a$, right-origin $\Last$).

\smallskip\noindent\textbf{Concurrent operations.}
\begin{itemize}
\item $o$: insert \texttt{"o"} with origin $\First$, right-origin $b$,
  id $(2,0)$.
\item $n$: insert \texttt{"n"} with origin $a$, right-origin $\Last$,
  id $(3,0)$.
\end{itemize}

\smallskip\noindent\textbf{Two delivery orders.}
\begin{center}
\begin{tabular}{ll}
  Order & Result \\
  \hline
  $o$ then $n$ & $[\,a,\, n,\, o,\, b\,]$ \\
  $n$ then $o$ & $[\,a,\, o,\, b,\, n\,]$
\end{tabular}
\end{center}

\smallskip\noindent\textbf{Root cause.}
In isolation each of the four items is perfectly well-formed: their origin
and right-origin fields name existing pointers and the algorithm accepts
them individually.  The anomaly is that insert operations are generated with
\emph{adjacent} neighbors at broadcast time; item $o$ was broadcast with
origin $\First$ and right-origin $b$ because, at the generating replica, no
item lay strictly between them, while item $a$ originated at a different
replica in which $o$ was not yet present.  Under causal delivery $o$ could
never see a state already containing $a$, so the configuration above is
\emph{unreachable}.  Unconditional commutativity, however, is stated as a
property of the pure effect function over \emph{all} states, and the
unreachable initial state above is enough to refute it.

\smallskip\noindent\textbf{Consequence.}
The standard convergence framework of Gomes et al.~\cite{Gomes2017}, which
requires unconditional commutativity, cannot be applied to Yjs.  Our
conditional commutativity framework resolves this by requiring commutativity
only on invariant-satisfying states that causal delivery actually produces.

\subsection{Proof Outline}

We prove convergence in two layers.  The first is a \emph{generic} argument
(\S\ref{sec:convergence}): starting from a single delivery list, we reorder
it by repeatedly swapping adjacent concurrent operations until it matches
any other delivery list with the same underlying set.  Each adjacent swap
must preserve the state, so this layer needs
\emph{conditional} concurrent commutativity, and, because a swapped
operation is applied to a state different from the one it was broadcast
against, a \emph{causal validity} transport property.  The second layer
(\S\ref{sec:order}--\ref{sec:algorithm}) establishes these two premises for
Yjs.  The Yjs-specific argument proceeds as follows.

Section~\ref{sec:order} fixes a strict total order $\YjsLt$ on Yjs items
(totality, transitivity, and asymmetry), defined on any item set satisfying
an \emph{item-set invariant} (origin ordering, origin reachability, and ID
uniqueness).  The role of this invariant is to carve out exactly the
configurations on which the order is well-behaved; the proof also depends
on a key auxiliary result, the \emph{no-cross-origin lemma}, which controls
how the origins of ordered items can relate.  The no-cross-origin lemma is
used in Section~\ref{sec:algorithm} to show that Yjs insert and delete
operations preserve the invariant together with pairwise sortedness under
$\YjsLt$.  In particular, Yjs insert is a non-trivial procedural loop, and
proving order preservation requires discovering a four-part loop invariant
for the scan.  Once preservation is established, conditional commutativity
is immediate: two sorted arrays with the same underlying set are equal
under any strict total order.

Finally, causal validity for Yjs reduces to showing that an item's
broadcast-time validity survives the causal reordering: the argument uses
invariant preservation from Section~\ref{sec:algorithm} together with
monotonicity of the items array to transport validity to any causally
compatible state.

%%
%% -------------------------------------------------------------------
\section{Total Order on Items}
\label{sec:order}
%%

This section defines the strict total order $\YjsLt$ on Yjs items that
underpins every subsequent argument.  The order is only well-behaved on item
sets that are structurally closed and satisfy a small invariant; we therefore
begin by fixing those conditions, then define $\YjsLt$, and finally establish
totality, transitivity, and asymmetry.  We close the section with a key
auxiliary lemma---the \emph{no-cross-origin lemma}---that will be used in
Section~\ref{sec:algorithm} to show that the Yjs insert algorithm terminates
in the correct position.

%% P1: Closed Item Set, Origin Reachability, Item-Set Invariant
Because $\YjsLt$ is defined by recursion on the origin and right-origin
components of items, the set of pointers we reason about must be closed under
those fields; otherwise a recursive call could land outside the set.

\begin{definition}[Closed item set (\leanref{IsClosedItemSet})]\label{def:closed-item-set}
A set $P$ of Yjs pointers is a \emph{closed item set} when the following
hold:
\begin{itemize}
\item $\First \in P$ and $\Last \in P$;
\item whenever $\langle x \rangle \in P$ we have
  $x.\mathit{origin} \in P$ and $x.\mathit{rightOrigin} \in P$.
\end{itemize}
\end{definition}

The invariant we impose on a closed item set restricts, in addition, how
reachability through origins can relate to $\YjsLt$.  Origin reachability is
the least relation that takes one step to either $\mathit{origin}$ or
$\mathit{rightOrigin}$, transitively closed.

\begin{definition}[Origin reachability (\leanref{OriginReachable})]\label{def:origin-reachability}
\emph{Origin reachability} is the least relation $\mathit{Reach}$ satisfying:
\begin{mathpar}
\inferrule*[right=Reach-Origin]
  {}
  {\mathit{Reach}(\langle o,r,\mathit{id},c\rangle,\; o)}
\and
\inferrule*[right=Reach-RightOrigin]
  {}
  {\mathit{Reach}(\langle o,r,\mathit{id},c\rangle,\; r)}
\and
\inferrule*[right=Reach-Trans]
  {\mathit{Reach}(x, y) \\ \mathit{Reach}(y, z)}
  {\mathit{Reach}(x, z)}
\end{mathpar}
\end{definition}

\begin{definition}[Item-set invariant (\leanref{ItemSetInvariant})]\label{def:item-set-invariant}
A closed item set $P$ satisfies the \emph{item-set invariant} when the
following hold:
\begin{itemize}
\item[(I1)] \textbf{Origin ordering.} For every $\langle o, r, \mathit{id}, c \rangle \in P$,
  we have $o \YjsLt r$.
\item[(I2)] \textbf{Reachability condition.} For every $x = \langle o, r, \mathit{id}, c
  \rangle \in P$ and every pointer $z$ reachable from $x$ via $\mathit{Reach}$,
  either $z \YjsLeq o$ or $r \YjsLeq z$.
\item[(I3)] \textbf{ID uniqueness.} No two distinct items in $P$ share the same
  identifier.
\end{itemize}
\end{definition}
Condition~I2 is the key structural property violated in the counterexample of
\S\ref{sec:overview}: item~$o$ has origin $\First$ and right-origin $b$, yet
the pointer $a$ is reachable from~$o$ via $b$'s origin, and
$\First \YjsLt a \YjsLt b$, so $a$ lies strictly between $\First$ and $b$,
violating the ``no intermediary reachable'' requirement.

%% P2: Yjs Order
With the invariant in place, we now define the Yjs order.  We will show that
on any closed item set satisfying the invariant, the defined relation is a
strict total order (Theorems~\ref{thm:totality}--\ref{thm:asymmetry}).  The
rules in Figure~\ref{fig:yjs-order} encode the Yjs tiebreaker: when two
items compete for the same interval, the one with the smaller origin (under
$\YjsLt$) wins; among items with equal origins, the one with the smaller
identifier wins.

\begin{figure*}[t]
\begin{mathpar}
\inferrule*[right=Lt-OrigOrd]
  {p_1 \OriginLt p_2}
  {p_1 \YjsLt p_2}
\and
\inferrule*[right=Lt-Origin]
  {p \YjsLeq x.\mathit{origin}}
  {p \YjsLt \langle x \rangle}
\and
\inferrule*[right=Lt-RightOrigin]
  {x.\mathit{rightOrigin} \YjsLeq p}
  {\langle x \rangle \YjsLt p}
\and
\inferrule*[right=Lt-Conflict]
  {x \mathrel{<_C} y}
  {\langle x \rangle \YjsLt \langle y \rangle}
\and
\inferrule*[right=Conflict-DiffOrigin]
  {y.\mathit{origin} \YjsLt x.\mathit{origin} \\
   \langle x \rangle \YjsLt y.\mathit{rightOrigin} \\
   x.\mathit{origin} \YjsLt \langle y \rangle \\
   \langle y \rangle \YjsLt x.\mathit{rightOrigin}}
  {x \mathrel{<_C} y}
\and
\inferrule*[right=Conflict-SameOrigin]
  {x.\mathit{origin} = y.\mathit{origin} \\
   \langle x \rangle \YjsLt y.\mathit{rightOrigin} \\
   \langle y \rangle \YjsLt x.\mathit{rightOrigin} \\
   x.\mathit{id} < y.\mathit{id}}
  {x \mathrel{<_C} y}
\end{mathpar}
\caption{The Yjs order $\YjsLt$ and the conflict order $\mathrel{<_C}$,
  defined by simultaneous induction.  $\YjsLeq$ is the reflexive closure of
  $\YjsLt$; $p \OriginLt q$ holds when $p = \First$ and $q \neq \First$, or
  $p \neq \Last$ and $q = \Last$.}
\label{fig:yjs-order}
\end{figure*}

\begin{definition}[Yjs order (\leanref{YjsLt}, \leanref{YjsLeq}, \leanref{ConflictLt})]\label{def:yjs-order}
The relations $\YjsLt$, $\YjsLeq$, and the \emph{conflict order}
$\mathrel{<_C}$ are defined by simultaneous induction via the rules of
Figure~\ref{fig:yjs-order}.
\end{definition}

\begin{theorem}[Totality (\leanref{yjs\_lt\_total})]
\label{thm:totality}
For any closed item set $P$ satisfying the item-set
invariant, and any $x, y \in P$, either $x \YjsLeq y$ or
$y \YjsLt x$.
\end{theorem}

\begin{theorem}[Transitivity (\leanref{yjs\_lt\_trans})]
\label{thm:transitivity}
For any closed item set $P$ satisfying the item-set invariant,
$\YjsLt$ is transitive on $P$.
\end{theorem}

\begin{theorem}[Asymmetry (\leanref{yjs\_lt\_asymm})]
\label{thm:asymmetry}
For any closed item set $P$ satisfying the item-set invariant,
$\YjsLt$ is asymmetric on $P$.
\end{theorem}

\noindent
All three proofs proceed by strong induction on the structural size of the
pointers involved ($|x| + |y|$ or $|x| + |y| + |z|$), where
$|\First| = |\Last| = 0$ and
$|\langle o, r, \mathit{id}, c \rangle| = |o| + |r| + 2$.
Together, these three properties establish that $\YjsLt$ is a strict total
order on the items of any $P$ satisfying the invariant.  Detailed proofs
appear in Appendix~\ref{app:order-proofs}.

%% P4: No-Cross-Origin
We conclude this section with a key lemma used in Section~\ref{sec:algorithm}
to prove that the Yjs insert scan terminates at a position respecting the
total order.

\begin{lemma}[No-Cross-Origin, \leanref{no\_cross\_origin}]
\label{lem:no-cross-origin}
Let $P$ be a closed item set satisfying the item-set invariant.
For any $x, y \in P$ with $x \YjsLt y$ (as items, i.e.\ $\langle x \rangle
\YjsLt \langle y \rangle$), either
$y.\mathit{origin} \YjsLeq x.\mathit{origin}$
or
$\langle x \rangle \YjsLeq y.\mathit{origin}$.
\end{lemma}

Intuitively, this says that if $x$ precedes $y$ in the order, then $y$'s
origin does not ``cross over'' $x$: it is either at or left of $x$'s origin,
or at or right of $x$ itself.  Figure~\ref{fig:no-cross-origin} illustrates
the two cases.

\begin{figure}[t]
\centering
\begin{tikzpicture}[>=Stealth, every node/.style={font=\small},
             item/.style={circle, draw, inner sep=1.5pt, minimum size=14pt},
             ptr/.style={rectangle, draw, inner sep=1.5pt, minimum size=12pt,
                         rounded corners=2pt},
             origin/.style={->, thick, bend right=60, densely dashed}]
%% Case (a): y_o <= x_o  - y's origin at or left of x's origin
\begin{scope}[local bounding box=caseA]
  \node[ptr]  (yoA) at (0.0,0) {$y_o$};
  \node[ptr]  (xoA) at (1.5,0) {$x_o$};
  \node[item] (xA)  at (3.0,0) {$x$};
  \node[item] (yA)  at (4.5,0) {$y$};
  %% origin reference arrows (dashed)
  \draw[origin] (xA) to (xoA);
  \draw[origin] (yA) to (yoA);
  %% horizontal brace for y_o <= x_o
  \draw[decorate, decoration={brace, amplitude=4pt, mirror}]
    ([yshift=-2pt]yoA.south) -- ([yshift=-2pt]xoA.south)
    node[midway, below=4pt, font=\scriptsize] {$y_o \YjsLeq x_o$};
\end{scope}
\node[above=10pt] at (caseA.north) {\textbf{Case (a):} $y_o \YjsLeq x_o$};

%% Case (b): x <= y_o  - x at or left of y's origin
\begin{scope}[local bounding box=caseB, shift={(0,-2.8)}]
  \node[ptr]  (xoB) at (0.0,0) {$x_o$};
  \node[item] (xB)  at (1.5,0) {$x$};
  \node[ptr]  (yoB) at (3.0,0) {$y_o$};
  \node[item] (yB)  at (4.5,0) {$y$};
  \draw[origin] (xB) to (xoB);
  \draw[origin] (yB) to (yoB);
  \draw[decorate, decoration={brace, amplitude=4pt, mirror}]
    ([yshift=-2pt]xB.south) -- ([yshift=-2pt]yoB.south)
    node[midway, below=4pt, font=\scriptsize] {$x \YjsLeq y_o$};
\end{scope}
\node[above=10pt] at (caseB.north) {\textbf{Case (b):} $x \YjsLeq y_o$};
\end{tikzpicture}
\caption{The No-Cross-Origin lemma (Lemma~\ref{lem:no-cross-origin}).
  Items are circles, pointers (origins) are rectangles, and dashed arrows
  indicate the $\mathit{origin}$ field of each item; positions on the line
  reflect the order $\YjsLt$.  Assuming $x \YjsLt y$, either (a)~$y$'s
  origin lies at or left of $x$'s origin or (b)~$x$ itself lies at or left
  of $y$'s origin; the ``forbidden'' configuration---$y_o$ strictly between
  $x_o$ and $x$---cannot arise.}
\label{fig:no-cross-origin}
\end{figure}

\noindent
The proof uses strong induction on $|x| + |y|$; the key case
(\textsc{Lt-RightOrigin}) relies on condition~I2 to chain origin
reachability.  See Appendix~\ref{app:order-proofs} for the full proof.

%%
%% -------------------------------------------------------------------
\section{Insert Algorithm and Commutativity}
\label{sec:algorithm}
%%

%% P1: Goal is commutativity
The goal of this section is to establish concurrent commutativity for Yjs,
that is, to show that concurrent operations commute when both are valid at
the state at which they are applied.  Three cases arise.  Delete-delete is
immediate because deletion only adds to the $\mathit{deletedIds}$ set and
the set union is commutative.  Insert-delete is immediate for a similar
reason: the items array and $\mathit{deletedIds}$ are independent
components of the state, so an insert and a delete touch disjoint data and
always commute.  The interesting case is insert-insert, to which the rest
of this section is devoted.  The strategy, foreshadowed in
\S\ref{sec:overview}, is that the Yjs insert operation preserves an
invariant that includes pairwise sortedness under $\YjsLt$: once
preservation is established, the two delivery orders produce sorted arrays
containing the same items, and a strict total order admits at most one such
arrangement, so the two results coincide.

%% P2: State Invariant
We begin by stating the invariant that states must satisfy throughout the
execution; it is the state-level counterpart of the item-set invariant of
\S\ref{sec:order}, together with a pairwise-sortedness condition on the
items array.

\begin{definition}[State invariant (\leanref{YjsArrInvariant})]
\label{def:state-inv}
A state $s = (\mathit{arr}, \mathit{del})$ satisfies the \emph{Yjs array
invariant} when the following hold:
\begin{enumerate}
\item $\mathit{ArrSet}(\mathit{arr})$ is a closed item set.
\item $\mathit{ArrSet}(\mathit{arr})$ satisfies the item-set invariant
  (I1--I3).
\item $\mathit{arr}$ is pairwise sorted by $\YjsLt$:
  for all $i < j$, $\mathit{arr}[i] \YjsLt \mathit{arr}[j]$.
\item All item identifiers in $\mathit{arr}$ are pairwise distinct.
\end{enumerate}
Here $\mathit{ArrSet}(\mathit{arr})$ is the item set induced by the array,
containing $\First$, $\Last$, and each item in $\mathit{arr}$.
\end{definition}

%% P5: Item Validity
The broadcast validity predicate (V) from Definition~\ref{def:operation-network}
requires each broadcast insert to be valid at the state at which it is
broadcast.  For Yjs we make that predicate concrete: an insert's item must
satisfy the per-item analogues of I1 and I2, evaluated in the item set
induced by the state to which the insert is being applied.

\begin{definition}[Item validity (\leanref{IsItemValid})]
\label{def:item-valid}
An item $x = \langle o, r, \mathit{id}, c \rangle$ is \emph{valid} in an
item set $P$ (containing $x$ together with its ancestors) when the following
hold:
\begin{enumerate}
\item $o \YjsLt r$ (origin strictly precedes right-origin);
\item for every pointer $z$ reachable from $x$ via $\mathit{Reach}$,
  either $z \YjsLeq o$ or $r \YjsLeq z$.
\end{enumerate}
An insert input $\mathit{inp}$ is valid at a state $s$, written
$\mathit{inp}.\mathit{isValid}(s)$, when the item
$\mathit{inp}.\mathit{toItem}(s)$ it resolves to is valid in
$\mathit{ArrSet}(s.\mathit{arr})$; this is the predicate used to instantiate
condition~(V).
\end{definition}

\noindent
Item validity is state-dependent: the same item may be valid at one state
but invalid at another.  Broadcast validity speaks of the state at which the
item is \emph{issued}, whereas the preservation and commutativity theorems
below need validity at the state to which the insert is \emph{applied}.  The
bridge between the two---sufficient under causal delivery---is deferred to
\S\ref{sec:convergence} as \emph{causal validity}.

%% P6: Insert Preservation
\begin{theorem}[Insert preservation (\leanref{YjsArrInvariant\_integrate})]
\label{thm:ins-pres}
If state $s$ satisfies the Yjs array invariant and $\mathit{inp}$ is an
insert input valid at $s$, then
$\effect(\mathsf{Insert}(\mathit{inp}), s)$ also satisfies the invariant.
\end{theorem}

\noindent
The analogous preservation for deletes is trivial, since deletion only
modifies $\mathit{deletedIds}$; we therefore focus on insert preservation.
The proof is a standard loop-invariant argument on
Algorithm~\ref{alg:integrate}: we give a four-part invariant for the scan,
show that each iteration preserves it, and show that the invariant implies
the desired post-condition at termination.

%% P3: Loop Invariant
\begin{definition}[Integration invariant (\leanref{IntegrationInvariant})]
\label{def:loop-inv}
During the insert scan, the \emph{integration invariant} at position~$i$
(with candidate destination~$\mathit{dest}$, flag~$\mathit{scanning}$, and
termination flag~$\mathit{done}$) asserts the following.
\begin{enumerate}
\item \textbf{Confirmed prefix} (\texttt{prefixLt}).
  For all $j < \mathit{dest}$, $\mathit{arr}[j] \YjsLt \mathit{newItem}$.
\item \textbf{Pending region} (\texttt{scannedRegion}).
  For all $\mathit{dest} \leq j < i$, one of the following holds:
  \begin{itemize}
  \item[(a)] $\mathit{arr}[j].\mathit{origin} = \mathit{newItem}.\mathit{origin}$
    and $\mathit{newItem}.\mathit{id} < \mathit{arr}[j].\mathit{id}$;
  \item[(b)] $\mathit{newItem}.\mathit{origin} \YjsLt
    \mathit{arr}[j].\mathit{origin}$.
  \end{itemize}
\item \textbf{Scanning flag} (\texttt{scanningOrigin}).
  $\mathit{scanning}$ is set exactly when
  $\mathit{arr}[\mathit{dest}].\mathit{origin} =
  \mathit{newItem}.\mathit{origin}$.
\item \textbf{Boundary ordering} (\texttt{doneLt}).
  If $\mathit{done}$, then $\mathit{newItem} \YjsLt \mathit{arr}[i]$.
\end{enumerate}
\end{definition}

\noindent
The pending region collects those array positions whose ordering against
$\mathit{newItem}$ the scan has not yet resolved.  Case~(a) applies when
the element shares $\mathit{newItem}$'s origin but has a larger id; deciding
it via \textsc{Conflict-SameOrigin} is blocked because
$\mathit{newItem} \YjsLt \mathit{arr}[j].\mathit{rightOrigin}$ has not yet
been established by the scan.  Case~(b) applies when the element has a
strictly greater origin; \textsc{Conflict-DiffOrigin} is blocked for the
same reason.  The \textbf{doneLt} clause records that, when the scan stops,
it has found an element strictly greater than $\mathit{newItem}$; the
pending propagation lemma below uses this information to promote every
pending element to a confirmed one, showing that $\mathit{dest}$ is the
smallest position at which $\mathit{newItem}$ can be inserted while
preserving pairwise sortedness.

\paragraph{Initialization.}
The scan starts at $i = \mathit{leftIdx} + 1$ with
$\mathit{dest} = \mathit{leftIdx} + 1$,
$\mathit{scanning} = \mathbf{false}$, and
$\mathit{done} = \mathbf{false}$.  The confirmed prefix contains all
elements before $\mathit{leftIdx} + 1$, each $\YjsLt \mathit{newItem}$
because $\mathit{newItem}.\mathit{origin}$ sits at position
$\mathit{leftIdx}$ and the array is sorted; the pending region and
scanning flag are vacuous, and boundary ordering holds vacuously.

\paragraph{Preservation (\leanref{loopInv\_preserve1}).}
Let $u = \mathit{newItem}$ and $o = \mathit{arr}[i]$ be the scanned
element.  The scan branches on the order between $u.\mathit{origin}$ and
$o.\mathit{origin}$, and, in the equal-origin subcase, on the id comparison
and on whether $o.\mathit{rightOrigin} = u.\mathit{origin}$:
\begin{itemize}
\item \emph{$o.\mathit{origin} \YjsLt u.\mathit{origin}$}: \textsc{break}
  with $\mathit{done}$ set.
\item \emph{$o.\mathit{origin} = u.\mathit{origin}$}:
  (i)~if $u.\mathit{id} > o.\mathit{id}$, advance $\mathit{dest}$ to
  $i+1$ and clear $\mathit{scanning}$;
  (ii)~else if $o.\mathit{rightOrigin} = u.\mathit{origin}$,
  \textsc{break};
  (iii)~else, set $\mathit{scanning}$ and keep the element pending.
\item \emph{$u.\mathit{origin} \YjsLt o.\mathit{origin}$}: keep the
  element pending (scanning flag unchanged).
\end{itemize}
The last branch splits further according to the scanning flag (advance
$\mathit{dest}$ when $\neg\mathit{scanning}$, otherwise only extend the
pending region), giving six sub-cases in total.  Preservation is discharged
uniformly: \textbf{doneLt} in the two \textsc{break} branches follows from
\textsc{Conflict-SameOrigin} or from no-cross-origin
(Lemma~\ref{lem:no-cross-origin}), which rules out the configuration in
which $o$ would lie at or left of $u.\mathit{origin}$ when
$o.\mathit{origin} \YjsLt u.\mathit{origin}$; \textbf{prefixLt} in the
advance branches follows from pending propagation applied to the current
region, combined with the new inequality produced by the branch; and
\textbf{scannedRegion} in the pending-extension branches follows directly
from the branch condition, without any appeal to no-cross-origin (this is
the reason for using $u.\mathit{origin} \YjsLt \mathit{arr}[j].\mathit{origin}$
in case~(2b) of the invariant).  The full case analysis is in
Appendix~\ref{app:preservation-cases}.

\paragraph{Postcondition.}
When the scan terminates the invariant gives a destination~$d$ and a loop
bound~$i$ such that $\mathit{arr}[j] \YjsLt u$ for all $j < d$ (confirmed
prefix) and that $u \YjsLt \mathit{arr}[i]$ (from \textbf{doneLt} in the
\textsc{break} case, or vacuously when the scan reaches
$\mathit{rightIdx}$).  The pending region $d \leq j < i$ must be collapsed
onto the confirmed prefix; this is precisely the role of the pending
propagation lemma.

\begin{lemma}[Pending propagation (\leanref{pending\_propagation})]
\label{lem:pending-prop}
Suppose the integration invariant holds at position~$i$ and
$u \YjsLt \mathit{arr}[i]$.  Then $u \YjsLt \mathit{arr}[j]$ for every
$j$ with $\mathit{dest} \leq j < i$.
\end{lemma}

\noindent\textit{Proof sketch.}
By strong induction on the size of $a_j := \mathit{arr}[j]$.  Let
$r_1 = u.\mathit{rightOrigin}$ and $r_2 = a_j.\mathit{rightOrigin}$.

\textbf{Case (a):} $a_j.\mathit{origin} = u.\mathit{origin}$ and
$u.\mathit{id} < a_j.\mathit{id}$.  We apply
\textsc{Conflict-SameOrigin}, which requires
$u \YjsLt r_2$ and $a_j \YjsLt r_1$.  The second holds because the
sorted invariant places $a_j$ strictly before the array position of
$r_1$, which the scan has not yet reached.  For the first, write
$r_2 = \mathit{arr}[k]$ for some index $k > j$.  If $k < i$ then $k$
itself lies in the pending region, so the inductive hypothesis gives
$u \YjsLt \mathit{arr}[k] = r_2$; if $i \leq k$ then
$u \YjsLt \mathit{arr}[i] \YjsLeq \mathit{arr}[k] = r_2$ by the scan
hypothesis combined with sortedness.  In either case,
\textsc{Conflict-SameOrigin} gives $u \YjsLt a_j$.

\textbf{Case (b):} $u.\mathit{origin} \YjsLt a_j.\mathit{origin}$.  Rule
\textsc{Lt-Origin} gives $u \YjsLt a_j$ once we have
$u \YjsLeq a_j.\mathit{origin}$.  When the scanning flag is set we have
$\mathit{arr}[\mathit{dest}].\mathit{origin} = u.\mathit{origin}$, so
$\mathit{arr}[\mathit{dest}] \YjsLt a_j$ by the sorted invariant, and
no-cross-origin (Lemma~\ref{lem:no-cross-origin}) upgrades this to
$\mathit{arr}[\mathit{dest}] \YjsLeq a_j.\mathit{origin}$, hence
$u.\mathit{origin} \YjsLt a_j.\mathit{origin}$ yields
$u \YjsLt a_j.\mathit{origin}$ by \textsc{Lt-Origin}.  The non-scanning
sub-case is similar, using the fact that $\mathit{dest}$ has already
absorbed the scan progress.

\smallskip
\noindent
With Lemma~\ref{lem:pending-prop}, insert preservation follows: at
termination the confirmed prefix gives $\mathit{arr}[d-1] \YjsLt u$,
pending propagation gives $u \YjsLt \mathit{arr}[d]$, so $d$ is the
unique position at which inserting $u$ preserves pairwise sortedness
(\leanref{YjsArrInvariant\_insert\-Idx\-If\-In\-Bounds}).  The other
three parts of the invariant transfer by routine calculation.

%% P7: Insert-Insert Commutativity
\begin{corollary}[Insert-insert commutativity, \leanref{integrate\_ok\_commutative}]
\label{thm:ins-ins}
Let $s$ satisfy the Yjs array invariant and let $a$, $b$ be two insert
inputs with $a.\mathit{clientId} \neq b.\mathit{clientId}$, both valid at
$s$.  Then $a$ remains valid after applying $b$ (and symmetrically), and
\[
  \effect(\mathsf{Insert}(a), \effect(\mathsf{Insert}(b), s)) =
  \effect(\mathsf{Insert}(b), \effect(\mathsf{Insert}(a), s)).
\]
\end{corollary}

\noindent\textit{Proof.}
Distinct client identifiers combined with clock monotonicity~(Y2) ensure
that the items induced by $a$ and $b$ have distinct identifiers, so
inserting one into a state satisfying the invariant keeps the other's
ancestors present and preserves the other's item validity.  By
Theorem~\ref{thm:ins-pres}, both orders of insertion produce
invariant-satisfying arrays, and those arrays contain the same underlying
item set (the items of $s$ together with the $a$- and $b$-items).  The
array invariant requires pairwise sortedness by $\YjsLt$ and $\YjsLt$ is a
strict total order on invariant-satisfying item sets
(Theorems~\ref{thm:totality}--\ref{thm:asymmetry}), so any two sorted
arrays with the same element set are identical
(\leanref{same\_yjs\_set\_unique}); this uniqueness is proved by induction
on list length using asymmetry to force the heads to agree.

\smallskip
\noindent
Corollary~\ref{thm:ins-ins} is precisely the concurrent-commutativity
statement we need, but its hypothesis appeals to item validity at $s$ for
both inserts, whereas the operation-network axiom~(V) only guarantees
validity at the state at which each insert was \emph{broadcast}.
Section~\ref{sec:convergence} closes this gap: causal delivery transports
broadcast-time validity to any replay state containing the insert's causal
predecessors.

%%
%% -------------------------------------------------------------------
\section{Convergence}
\label{sec:convergence}
%%

%% P1: Generic argument abstract from network model
The convergence proof proceeds in two layers.  The \emph{generic} layer of
\S\ref{subsec:generic} works entirely with abstract operation lists carrying
a happens-before relation~$\hb$; it culminates in a convergence theorem
(Theorem~\ref{thm:generic-conv}) whose hypotheses are that the two delivery
lists are $\hb$-consistent, $\hb$-closed, duplicate-free, satisfy a
\emph{conditional commutativity} condition, and admit a \emph{causal
validity} transport.  The \emph{Yjs} layer of \S\ref{subsec:yjs-inst}
instantiates this theorem: the first three premises follow from the causal
network axioms alone, while conditional commutativity and causal validity
are the parts specific to Yjs and build directly on the results of
\S\ref{sec:algorithm}.

\subsection{Generic Convergence}
\label{subsec:generic}

We first fix the language in which the generic theorem is stated.  Two
conditions on an operation list carry the causality information a
convergence argument needs: the list must be consistent with the
happens-before relation (no later element $\hb$-precedes an earlier one)
and closed under $\hb$-predecessors (nothing causally necessary is
missing).

\begin{definition}[$\hb$-consistency]
\label{def:hb-consistent}
A list $L$ of operations is \emph{$\hb$-consistent}
($\mathit{hbConsistent}(L)$) when it satisfies the following inductive
predicate:
\begin{mathpar}
\inferrule*[right=HbC-Nil]
  {}
  {\mathit{hbConsistent}([])}
\and
\inferrule*[right=HbC-Cons]
  {\mathit{hbConsistent}(L) \\
   \forall b \in L.\; \neg(b \hble a)}
  {\mathit{hbConsistent}(a :: L)}
\end{mathpar}
Equivalently, no element is $\hb$-preceded by a later element in the list.
A list $L$ is \emph{$\hb$-closed} when, for every $a \in L$ and every $b$
with $b \hb a$, we have $b \in L$.
\end{definition}

The convergence argument reorders operations by adjacent swaps, so we need
to know that swapping two concurrent operations preserves the conditions
above.  This is the content of the swap lemmas below.

\begin{lemma}[Swap preserves consistency and closure,
  \leanref{hbConsistent\_swap}, \leanref{hbClosed\_swap}]
\label{lem:swap}
Suppose $L = L_0 \mathbin{+\!\!+} [a, b] \mathbin{+\!\!+} L_1$ and $a$ and
$b$ are concurrent ($a \not\hble b$ and $b \not\hble a$).  Then
$L_0 \mathbin{+\!\!+} [b, a] \mathbin{+\!\!+} L_1$ is $\hb$-consistent whenever
$L$ is, and likewise for $\hb$-closure.
\end{lemma}

\noindent\textit{Proof sketch.}
Because $a$ and $b$ are concurrent, every element $\hb$-before $a$ is
either $\hb$-before $b$ or already to the left of $[a,b]$, and
symmetrically.  Thus neither condition on $L$ can be violated by swapping
the adjacent pair.

Commutativity of concurrent operations is what turns a swap into a
state-preserving transformation.  We state it in the general,
state-dependent form that our proof actually requires.

\begin{definition}[Conditional concurrent commutativity
  (\leanref{concurrent\_commutative})]
\label{def:conditional-comm}
A list of operations $L$ satisfies \emph{conditional concurrent
commutativity} (with respect to a state invariant $\Phi$ and validity
predicate $\mathit{isValid}$) when, for every concurrent pair $a, b \in L$
and every state $s$ with $\Phi(s)$,
$\mathit{isValid}(s, a)$, and $\mathit{isValid}(s, b)$,
$\effect(a, \effect(b, s)) = \effect(b, \effect(a, s))$.
\end{definition}

Conditional commutativity lets us swap adjacent concurrent operations at a
\emph{specific} state.  Reordering a delivery list by repeated adjacent
swaps, on the other hand, applies each operation to many different states,
none of which is the state at which the operation was originally broadcast.
To justify these applications we need a \emph{causal validity} condition:
the insert must remain valid at every state produced by an
$\hb$-consistent, $\hb$-closed replay that already contains its
$\hb$-predecessors.

\begin{definition}[Causally validatable operation
  (\leanref{CausallyValidatableOperation})]
\label{def:causally-valid}
Let $(\mathit{St}, \init, \effect)$ be an operation structure on a message
type~$A$ equipped with a happens-before relation~$\hb$ and a set
$S \subseteq A$ of source messages.  The structure is \emph{causally
validatable} with respect to $S$ when there exist a validity predicate
$\mathit{isValid}(s, a)$ and a state invariant $\Phi(s)$ such that:
\begin{enumerate}
\item $\Phi(\init)$ holds, and whenever $\Phi(s)$,
  $\mathit{isValid}(s, a)$, and $\effect(a, s) = s'$, we have $\Phi(s')$.
\item \emph{Causal validity.}
  For every $a \in S$, every $\hb$-consistent, $\hb$-closed, duplicate-free
  list $L$ that contains every $\hb$-predecessor of $a$, and every state
  $s$ with $\interpOps(L, \init) = s$, the predicate
  $\mathit{isValid}(s, a)$ holds.
\end{enumerate}
\end{definition}

\noindent
Informally, causal validity says that the validity of $a$ depends only on
the presence of its $\hb$-predecessors in the replay, not on the order in
which they are applied or on the state at which $a$ was originally
broadcast.  This \emph{downward-closed} character is exactly what lets us
apply $a$ to any reordered state produced by the swap argument: so long as
the replay has reached a state whose list contains the causal predecessors,
$a$ is valid there.

\begin{theorem}[Generic convergence,
  \leanref{hb\_consistent\_effect\_convergent}]
\label{thm:generic-conv}
Suppose the operation semantics is causally validatable with respect to a
source set $S$ (Definition~\ref{def:causally-valid}), and let $L_0$ and
$L_1$ be two lists with every element of $L_0 \cup L_1$ in $S$, both
$\hb$-consistent, both $\hb$-closed, both duplicate-free, and with $L_0$
satisfying conditional concurrent commutativity.  If $L_0$ and $L_1$
contain the same elements, then
$\interpOps(L_0, \init) = \interpOps(L_1, \init)$.
\end{theorem}

\noindent\textit{Proof sketch.}
By reverse induction on $L_1$; write $L_1 = L_1' \mathbin{+\!\!+} [b]$.
Locate $b$ in $L_0$ and, by repeated application of
Lemma~\ref{lem:swap}, move $b$ to the end of $L_0$ by swapping it past
adjacent concurrent operations one at a time.  Each swap preserves the
computed state by conditional commutativity; causal validity is needed at
each swap because the operation applied on the right of the pair sees a
different state from the one at which it was originally broadcast, and
Definition~\ref{def:causally-valid} guarantees that it is nevertheless
valid there.

\subsection{Yjs Instantiation}
\label{subsec:yjs-inst}

To apply Theorem~\ref{thm:generic-conv} to Yjs, we take $S$ to be the set
of messages broadcast by some node, $\Phi$ to be the Yjs array invariant
(Definition~\ref{def:state-inv}), and $\mathit{isValid}$ to be the
broadcast-validity predicate of the operation network
(Definition~\ref{def:item-valid}).  Under these choices, the
state-invariant preservation clause of
Definition~\ref{def:causally-valid} follows from
Theorem~\ref{thm:ins-pres}, and the two substantial premises---causal
validity and conditional concurrent commutativity---are established by the
two lemmas below.

\begin{lemma}[Yjs causal validity (\leanref{isValidState\_insert\_causally})]
\label{lem:yjs-causal-valid}
The Yjs operation network is causally validatable with source set the
messages broadcast by some node.
\end{lemma}

\noindent\textit{Proof sketch.}
Broadcast validity~(V) gives us, for each source $a$, a state $s^\star$ at
which $a$'s item is valid: the local state at the broadcasting replica
just before the broadcast.  Let $L$ be any $\hb$-consistent, $\hb$-closed,
duplicate-free list containing every $\hb$-predecessor of $a$, and let $s
= \interpOps(L, \init)$.  We transport item validity from $s^\star$ to $s$
in two steps.  First, every item present at $s^\star$ is also present at
$s$, because the $\hb$-predecessors of $a$ (which by broadcast validity
populate $s^\star$) are all contained in $L$ by assumption, and the items
array is monotone under insert/delete.  Second, item validity depends only
on membership of the relevant origin chain in the item set, not on
positions within the sorted array; it is therefore preserved by any
superset that still satisfies the array invariant, and
Theorem~\ref{thm:ins-pres} ensures that $s$ does satisfy the invariant.

\begin{lemma}[Yjs concurrent commutativity,
  \leanref{YjsOperationNetwork\_concurrentCommutative}]
\label{thm:yjs-comm}
For any Yjs operation network, the list of delivered operations at any
node satisfies conditional concurrent commutativity with respect to the
Yjs array invariant.
\end{lemma}

\noindent\textit{Proof sketch.}
A case analysis on the types of the concurrent pair.  The
delete-delete and insert-delete cases are immediate from the structure of
the state, as discussed at the start of \S\ref{sec:algorithm}.  The
insert-insert case combines two ingredients: distinct client identifiers
(from the distinct sources~(Y1)) and the fact that the two items are not
origin-reachable from each other, which is derivable from causal validity
(Lemma~\ref{lem:yjs-causal-valid}) by an argument that any item
origin-reachable from the first one must already be present in the
earlier broadcast state and so cannot be the second one; the argument is
given in Appendix~\ref{app:not-reach}.  With these inputs,
Corollary~\ref{thm:ins-ins} concludes.

Finally, the three \emph{structural} premises---$\hb$-consistency,
$\hb$-closure, and duplicate-freedom of
$\Delivered{i}(H_i)$---are properties of any causal network and do not
involve the Yjs-specific layer at all.

\begin{lemma}[Structural properties of delivered messages
  (\leanref{hb\_consistent\_local\_history},
   \leanref{toDeliverMessages\_hbClosed},
   \leanref{toDeliverMessages\_IdNoDup})]
\label{lem:structural-delivered}
For every node $i$ of a (causal) operation network,
$\Delivered{i}(H_i)$ is $\hb$-consistent, $\hb$-closed, and duplicate-free.
\end{lemma}

\noindent\textit{Proof sketch.}
(1)~$\hb$-consistency holds because causal delivery forces any $\hb$-predecessor
of a delivered message to be delivered before it, so no later element in
the delivery list $\hb$-precedes an earlier one.
(2)~$\hb$-closure holds because the same axiom ensures every
$\hb$-predecessor of a delivered message is itself delivered at the same
node.
(3)~Duplicate-freedom follows from the network's message-ID uniqueness
axiom~(N3).  Further details are in Appendix~\ref{app:structural-delivered}.

%% P7: Yjs Convergence
\begin{theorem}[Yjs Convergence (\leanref{YjsOperationNetwork\_converge'})]
\label{thm:yjs-convergence}
For any Yjs operation network, if nodes $i$ and $j$
deliver the same set of operations
($\Delivered{i}(H_i) \approx \Delivered{j}(H_j)$), then
$\interpOps(\Delivered{i}(H_i), \init) =
 \interpOps(\Delivered{j}(H_j), \init)$.
\end{theorem}

\noindent\textit{Proof.}
Apply Theorem~\ref{thm:generic-conv} with
$L_0 = \Delivered{i}(H_i)$ and $L_1 = \Delivered{j}(H_j)$; the required
hypotheses are discharged by the lemmas of this subsection.

%%
%% -------------------------------------------------------------------
\section{Related Work}
\label{sec:related}
%%

%% P1: Prior CRDT verification — methodological comparison
The closest related work is Gomes et al.~\cite{Gomes2017,GomesRGA2017}, who
developed a framework for verifying strong eventual consistency of CRDTs in
Isabelle/HOL and applied it to the RGA algorithm.
Their framework requires \emph{unconditional} concurrent commutativity:
operations must commute on \emph{all} states, not just reachable ones.
This assumption holds for RGA, where concurrent insertions are ordered by
globally unique timestamps without reference to the current state, but fails
for Yjs, as our counterexample demonstrates.
Our conditional commutativity framework strictly generalizes theirs:
unconditional commutativity is the special case where the invariant $\Phi$ is
the trivially true predicate.  The key methodological contribution is showing
that restricting commutativity to invariant-satisfying reachable states,
combined with the causal validity condition, suffices for convergence.
This framework is not specific to Yjs and can be applied to any CRDT whose
operations commute only on reachable states.

Zeller et al.~\cite{Zeller2014} formalize CRDTs in Coq, and
Lim and Kim~\cite{Lim2019} develop a mechanized CRDT theory; neither addresses
the Yjs/YATA family.  Other sequence CRDTs have been verified
in restricted settings~\cite{Netzell2020}.

Weidner et al.~\cite{Weidner2023} design the \emph{Fugue} sequence CRDT and,
in its appendix, provide a by-hand proof for a variant of the Yjs
integration algorithm.  Their development does not isolate conditional
commutativity as an explicit property, nor does it formulate a causal
validity condition; instead, the proof implicitly relies on the fact that
concurrent operations originating at causally adjacent states preserve an
invariant analogous to ours, which is precisely the content of conditional
commutativity.  To the best of our understanding of the by-hand
appendix, that reliance surfaces in the step showing that the integration
scan lands at the unique sorted position, where the argument appeals to
well-formedness of the current document---a property that holds only for
reachable states.  Making this reliance explicit and machine-checked is one
of the contributions of the present paper.

%% P2: YATA paper and the bug
The YATA paper~\cite{Nicolaescu2016} introduces the algorithm that Yjs
implements, with proofs sketched for a subset of the required properties.
The proofs are not machine-checked, and a bug was subsequently discovered: the
original tiebreaker permitted non-convergent behavior, requiring a fix in the
Yjs implementation.  Our work provides the first complete machine-checked proof
of the corrected algorithm.

%% P3: Other sequence CRDTs
Other sequence CRDTs such as RGA~\cite{Roh2011} and Logoot~\cite{Weiss2010}
achieve unconditional commutativity by construction: RGA uses unique timestamps,
Logoot uses dense position identifiers.  Yjs's origin/right-origin encoding
creates structural dependencies on the existing document, which is the source
of both its expressiveness and the need for conditional commutativity.
The Attiya et al.~\cite{Attiya2016} specification work captures semantic
requirements for sequence CRDTs but does not mechanize proofs.

%%
%% -------------------------------------------------------------------
\section{Discussion and Conclusion}
\label{sec:conclusion}
%%

%% P1: Summary
We have introduced a conditional commutativity convergence framework that
generalizes the standard unconditional commutativity approach, and instantiated
it for Yjs to obtain the first complete machine-checked proof of strong
eventual consistency for the Yjs core CRDT algorithm.
The main theorem (\leanref{YjsOperationNetwork\_converge'}) is fully
machine-checked in Lean~4 with no axioms beyond those of its type theory.

%% P2: Conditional commutativity as a reusable framework
The conditional commutativity framework is independent of Yjs: any CRDT whose
operations commute on invariant-satisfying reachable states---rather than on
all states---fits naturally into our generic theorem
(Theorem~\ref{thm:generic-conv}).  We expect this framework to be applicable
to other CRDTs with state-dependent conflict resolution.  Whether there is
a CRDT other than Yjs that genuinely needs the \emph{causal validity}
condition (as opposed to conditional commutativity alone) is an open
question we leave for future work.

%% P3: Limitations and future work
The present formalization models the core CRDT algorithm with insert and
delete operations.  We do not model garbage collection, undo/redo,
subdocuments, or the awareness protocol.  The artifact contains an
incomplete indirect variant where items reference origins by identifier.
Extending the proof to these features is future work.

\bibliographystyle{ACM-Reference-Format}
\bibliography{refs}

\appendix

%%
%% -------------------------------------------------------------------
\section{Order-Theoretic Proof Sketches}
\label{app:order-proofs}
%%

This appendix gives detailed proof sketches for the totality, transitivity,
and asymmetry theorems of \S\ref{sec:order}.

\subsection{Totality}

\noindent\textit{Proof sketch} (Theorem~\ref{thm:totality}).
By strong induction on $|x| + |y|$.
The cases involving $\First$ or $\Last$ are immediate
by \textsc{Lt-OrigOrd}.  When both are items
$x = \langle x_o, x_r, \mathit{id}_x, c_x \rangle$ and
$y = \langle y_o, y_r, \mathit{id}_y, c_y \rangle$,
the proof attempts four recursive comparisons on
interval boundaries, each well-founded because
$|x_o|, |x_r| < |x|$ and $|y_o|, |y_r| < |y|$:
\begin{enumerate}
\item $x$ vs.\ $y_o$: if $x \YjsLeq y_o$, then
  $x \YjsLt y$ by \textsc{Lt-Origin}.
\item $y_r$ vs.\ $x$: if $y_r \YjsLeq x$, then
  $y \YjsLt x$ by \textsc{Lt-RightOrigin}.
\item $y$ vs.\ $x_o$: if $y \YjsLeq x_o$, then
  $y \YjsLt x$ by \textsc{Lt-Origin}.
\item $x_r$ vs.\ $y$: if $x_r \YjsLeq y$, then
  $x \YjsLt y$ by \textsc{Lt-RightOrigin}.
\end{enumerate}
If all four fail (i.e.\ $y_o \YjsLt x$, $x \YjsLt y_r$,
$x_o \YjsLt y$, $y \YjsLt x_r$), then $x$ and $y$
compete for the same interval, and a fifth comparison
on the origins decides via the conflict rules:
if $x_o \neq y_o$, \textsc{Conflict-DiffOrigin}
applies; if $x_o = y_o$, the ID comparison
$\mathit{id}_x$ vs.\ $\mathit{id}_y$ resolves the
tie via \textsc{Conflict-SameOrigin}; if
$\mathit{id}_x = \mathit{id}_y$, condition~I3
(ID uniqueness) forces $x = y$.

\subsection{Transitivity}

\begin{lemma}[Conflict transitivity
  (\leanref{conflict\_lt\_trans})]
\label{lem:conflict-lt-trans}
If\/ $\mathit{ConflictLt}'(x,y)$ and\/
$\mathit{ConflictLt}'(y,z)$, then $x \YjsLt z$,
assuming the transitivity IH holds for all triples
with total size smaller than $|x|+|y|+|z|$.
\end{lemma}
\noindent\textit{Proof sketch.}
From the two conflict derivations, the IH yields
$x_o \YjsLt z$ (since $|x_o| < |x|$) and
$x \YjsLt z_r$ (since $|z_r| < |z|$).
Totality then tests whether $x_r \YjsLeq z$ or
$x \YjsLeq z_o$; if either holds,
\textsc{Lt-RightOrigin} or \textsc{Lt-Origin} closes
the goal.  When both fail, $x$ and $z$ compete for
the same interval, and the four combinations of
\textsc{Conflict-DiffOrigin}/\textsc{Conflict-SameOrigin}
resolve via the conflict rules (in the
\textsc{DiffOrigin}~$\times$~\textsc{DiffOrigin} case,
a further IH on the origins produces $z_o \YjsLt x_o$;
the remaining three cases apply a conflict rule
directly).

\medskip
\noindent\textit{Proof sketch} (Theorem~\ref{thm:transitivity}).
By strong induction on $|x| + |y| + |z|$.
A case-analysis lemma (\leanref{yjs\_lt\_cases})
decomposes each $\YjsLt$ derivation into
\textsc{Lt-RightOrigin}, \textsc{Lt-Origin}, or
\textsc{Lt-Conflict}.
The proof handles all nine combinations of
derivation rules for $x \YjsLt y$ and $y \YjsLt z$.

\smallskip\noindent\textbf{$x \YjsLt y$ by \textsc{Lt-RightOrigin}
($x_r \YjsLeq y$):}
This case is uniform in $y \YjsLt z$.
If $x_r = y$, substitute directly.
If $x_r \YjsLt y$, the IH on $(x_r, y, z)$
applies because $|x_r| < |x|$;
then \textsc{Lt-RightOrigin} gives $x \YjsLt z$.

\smallskip\noindent\textbf{$x \YjsLt y$ by \textsc{Lt-Origin}
($x \YjsLeq y_o$):}
We sub-case on $y \YjsLt z$.

\begin{itemize}
\item \emph{$y \YjsLt z$ by \textsc{Lt-RightOrigin}
  ($y_r \YjsLeq z$)}:
  By I1, $y_o \YjsLt y_r$.  We must chain
  $x \YjsLeq y_o \YjsLt y_r \YjsLeq z$,
  but transitivity is not yet established.
  The key observation is that the IH is on
  $|x|+|y|+|z|$, and since
  $|y_o|+|y_r|+2 = |y|$ we have
  $|x|+|y_o|+|y_r| < |x|+|y| \le |x|+|y|+|z|$,
  so the IH applies to the triple $(x, y_o, y_r)$
  to give $x \YjsLt y_r$.  Then
  $x \YjsLt z$ follows from the IH on
  $(x, y_r, z)$ since $|y_r| < |y|$.

\item \emph{$y \YjsLt z$ by \textsc{Lt-Origin}
  ($y \YjsLeq z_o$)}:
  It suffices to show $x \YjsLt z_o$
  (then \textsc{Lt-Origin} gives $x \YjsLt z$).
  If $y = z_o$, we have $x \YjsLt y = z_o$.
  If $y \YjsLt z_o$, the IH on $(x, y, z_o)$
  applies because $|z_o| < |z|$.

\item \emph{$y \YjsLt z$ by \textsc{Lt-Conflict}}:
  A helper lemma extracts
  $y_o \YjsLt z$ from any conflict derivation
  of $y \YjsLt z$.  Then
  $x \YjsLeq y_o \YjsLt z$;
  the IH on $(x, y_o, z)$ applies because
  $|y_o| < |y|$.
\end{itemize}

\noindent\textbf{$x \YjsLt y$ by \textsc{Lt-Conflict}:}
A helper lemma extracts $x \YjsLt y_r$ from any
conflict derivation of $x \YjsLt y$.
When $y \YjsLt z$ by \textsc{Lt-RightOrigin}
($y_r \YjsLeq z$), the IH on $(x, y_r, z)$ applies
since $|y_r| < |y|$.
When $y \YjsLt z$ by \textsc{Lt-Origin}
($y \YjsLeq z_o$), we reduce to showing
$x \YjsLt z_o$ and apply the IH on $(x, y, z_o)$
since $|z_o| < |z|$.
When both use \textsc{Lt-Conflict}, we apply
Lemma~\ref{lem:conflict-lt-trans}.

\subsection{Asymmetry}

\noindent\textit{Proof sketch} (Theorem~\ref{thm:asymmetry}).
By strong induction on $|x| + |y|$, assuming $x \YjsLt y$
and $y \YjsLt x$ for contradiction.
The proof decomposes each derivation via
\leanref{yjs\_lt\_cases}.
If $x \YjsLt y$ was derived by \textsc{Lt-RightOrigin}
($x_r \YjsLeq y$), a dedicated reduction lemma
shows there exist $x', y' \in P$ with
$|x'| + |y'| < |x| + |y|$ and $x' \YjsLt y'$,
$y' \YjsLt x'$, contradicting the IH.
The case for \textsc{Lt-Origin} is symmetric.
When both sides use \textsc{Lt-Conflict},
a third reduction lemma
(\leanref{yjs\_lt\_conflict\_lt\_decreases})
similarly produces a smaller contradictory pair.

\subsection{No-Cross-Origin}

\noindent\textit{Proof sketch} (Lemma~\ref{lem:no-cross-origin}).
By strong induction on $|x| + |y|$.
The \textsc{Lt-Origin} case ($x \YjsLeq y_o$) yields
the right disjunct directly.
The \textsc{Lt-Conflict} case reads off
$y_o \YjsLt x_o$ from the conflict premises and
yields the left disjunct.
The key case is \textsc{Lt-RightOrigin}
($x_r \YjsLeq y$).  If $x_r = y$ then condition~I2
(origin reachability) gives $y_o \YjsLeq x_o$
directly.  If $x_r \YjsLt y$, the IH applies
(since $|x_r| < |x|$) to obtain either
$y_o \YjsLeq x_r.\mathit{origin}$ or
$x_r \YjsLeq y_o$; in the former branch, I2 again
chains $x_r.\mathit{origin} \YjsLeq x_o$ to produce
$y_o \YjsLeq x_o$.

%%
%% -------------------------------------------------------------------
\section{Integration Preservation: Detailed Cases}
\label{app:preservation-cases}
%%

Let $u = \mathit{newItem}$ and $o = \mathit{arr}[i]$ be the scanned element.

\begin{itemize}
\item \emph{$o.\mathit{origin} < u.\mathit{origin}$} ---
  \textsc{break} ($\mathit{done} \leftarrow \mathit{true}$).
  Obligation: \textbf{doneLt} ($u \YjsLt o$).
  By totality either $u \YjsLt o$ or $o \YjsLeq u$.
  If $o \YjsLt u$, Lemma~\ref{lem:no-cross-origin} yields
  $u.\mathit{origin} \YjsLeq o.\mathit{origin}$ or
  $o \YjsLeq u.\mathit{origin}$;
  the first contradicts $o.\mathit{origin} < u.\mathit{origin}$,
  the second contradicts $o$'s position past~$u.\mathit{origin}$
  in the sorted array.

\item \emph{$o.\mathit{origin} = u.\mathit{origin}$,
  $o.\mathit{id} < u.\mathit{id}$} ---
  $\mathit{dest} \leftarrow i{+}1$,
  $\mathit{scanning} \leftarrow \mathit{false}$.
  Obligation: \textbf{prefixLt} for the expanded
  prefix~$[0, i{+}1)$.
  For each $j$ in $[\mathit{dest}, i]$:
  $\mathit{arr}[j] \YjsLeq o$ holds by the sorted array, and
  $o \YjsLt u$ by \textsc{Conflict-SameOrigin}
  (same origin, smaller ID).
  Transitivity gives $\mathit{arr}[j] \YjsLt u$.

\item \emph{Same origin, $o.\mathit{id} \geq u.\mathit{id}$,
  same right-origin} ---
  \textsc{break} ($\mathit{done} \leftarrow \mathit{true}$).
  Obligation: \textbf{doneLt} ($u \YjsLt o$).
  Both items share origin and right-origin with
  $u.\mathit{id} \leq o.\mathit{id}$;
  \textsc{Conflict-SameOrigin} gives $u \YjsLt o$.

\item \emph{Same origin, $o.\mathit{id} \geq u.\mathit{id}$,
  different right-origin} ---
  $\mathit{scanning} \leftarrow \mathit{true}$.
  Obligation: \textbf{scannedRegion} for~$o$.
  Condition~(2a) holds:
  $o.\mathit{origin} = u.\mathit{origin}$
  and $u.\mathit{id} < o.\mathit{id}$
  (by unique-ID distinctness).

\item \emph{$o.\mathit{origin} > u.\mathit{origin}$,
  $\neg\mathit{scanning}$} ---
  $\mathit{dest} \leftarrow i{+}1$.
  Obligation: \textbf{prefixLt} for the expanded prefix.
  When $\mathit{scanning}$ is false, $\mathit{dest}$
  equals the current scan position, so every element
  in the old pending region was already confirmed.
  The origin of~$o$ lies at some array index
  $k < \mathit{dest}$, so $o.\mathit{origin} \YjsLt u$
  by the prefix condition.
  Combined with item validity ($o$ lies between
  $u.\mathit{origin}$ and $u.\mathit{rightOrigin}$),
  we obtain $o \YjsLt u$; transitivity
  extends to all $j \leq i$.

\item \emph{$o.\mathit{origin} > u.\mathit{origin}$,
  $\mathit{scanning}$} ---
  continue; extend pending region.
  Obligation: \textbf{scannedRegion},
  condition~(2b), which asks for
  $u.\mathit{origin} \YjsLt o.\mathit{origin}$.
  This is the scan hypothesis directly.
\end{itemize}

%%
%% -------------------------------------------------------------------
\section{Concurrent Inserts Are Not Origin-Reachable}
\label{app:not-reach}
%%

\begin{lemma}[Concurrent inserts are not origin-reachable
  (\leanref{concurrent\_insert\_not\_reachable})]
\label{lem:not-reach}
Let $a$, $b$ be two insert inputs, both valid at a state~$s$ satisfying
the array invariant.  Then neither of the induced items is
origin-reachable from the other.
\end{lemma}

\noindent\textit{Proof sketch.}
Suppose for contradiction that the item induced by $b$ were
origin-reachable from the item induced by $a$.  Because $a$'s origin and
right-origin already exist in~$s$ (they resolve against
$s.\mathit{items}$), every origin-reachable step from $a$'s item stays
inside $\mathit{ArrSet}(s.\mathit{arr})$
(\leanref{toItem\_reachable\_mem}).  Hence the item induced by~$b$
already sits in~$s$, so its identifier is present.  But $b$'s insert
only succeeds when the clock-safety check rules out an identifier clash
(\leanref{isClockSafe\_false\_of\_mem}), contradicting the assumption
that applying $b$ at the state in which $b$ was broadcast succeeded.

%%
%% -------------------------------------------------------------------
\section{Structural Properties of Delivered Messages}
\label{app:structural-delivered}
%%

\noindent\textit{Proof of Lemma~\ref{lem:structural-delivered}.}
Fix a node~$i$ and write $L = \Delivered{i}(H_i)$.

\smallskip\noindent\textbf{$\hb$-consistency.}
By induction on $L$.  Suppose $L = a :: L'$ and, for contradiction, some
later $b \in L'$ satisfies $b \hble a$.  By definition of~$\hb$ this
means $b \hb a$ or $b = a$; the latter is excluded because events in
$H_i$ are duplicate-free.  The causal delivery axiom then forces
$\Deliver(b) <_i \Deliver(a)$ in~$H_i$, hence $b$ appears before $a$ in
$L$---contradicting $b \in L'$.

\smallskip\noindent\textbf{$\hb$-closure.}
For $a \in L$ and any $b \hb a$, the causal delivery axiom requires
$\Deliver(b) <_i \Deliver(a)$ in~$H_i$, so $b \in L$.

\smallskip\noindent\textbf{Duplicate-freedom.}
Two distinct occurrences of the same message in~$L$ would correspond to
two distinct $\Deliver$ events in~$H_i$ carrying the same $\mathit{msgId}$;
by axiom~(N3) together with (N1)/(N2) both trace back to the same
broadcast event, contradicting event duplicate-freedom within~$H_i$.

\end{document}
